---
title: "Wind Potential"
author: "Sam Campbell & John Rooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Install packages, message=FALSE, warning=FALSE}
#Import the tidyverse library 
library(tidyverse, quietly = TRUE)
library(lubridate)

#install.packages('sf')
library(sf)
#install.packages('leaflet')
library(leaflet)
#install.packages('mapview')
library(mapview); mapviewOptions(fgb = FALSE)

#Disable on-the-fly projections
sf::sf_use_s2(FALSE)

```


```{r join.attributes.to.spatial features}
#Read the 2018 SVI county-level dataset for NC
svi2018_nc_raw <- read.csv(
  'https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/NorthCarolina_COUNTY.csv',
  colClasses = c('FIPS' = 'factor')) %>% 
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV, E_MINRTY)

#Check structure
str(svi2018_nc_raw)

#Join the SVI attributes to the county spatial features
counties_sf_join <-  merge(x = counties_sf,
                           y = svi2018_nc_raw, 
                           by.x = "GEOID", 
                           by.y = "FIPS" )

#Tidyverse version of the join
counties_sf_join <- counties_sf %>% 
  left_join(svi2018_nc_raw, by = c("GEOID" = "FIPS") )

#View with mapview
mapview(counties_sf_join, 
        zcol = 'E_POV', 
        col.regions = brewer.pal(2, 'RdBu')) + 
  mapview(epa_pm25_sites_sf, cex = 'maxPM')

#view with ggplot
ggplot() + 
  geom_sf(data=counties_sf_join,aes(fill = E_POV),alpha=0.3) + 
  scale_fill_gradient2(low="red",high="blue",midpoint = 60000) + 
  geom_sf(data=epa_pm25_sites_sf)

```


```{r}
#Compute HUC8 wind energy
wind_raw <- read.csv('https://raw.githubusercontent.com/ENV859/EnviroAtlasData/main/Wind_Energy.csv',
  colClasses = c('HUC_12' = 'factor')) %>%  
  mutate(HUC_8 = substr(HUC_12,1,8)) %>% 
  group_by(HUC_8) %>% 
  summarize(SumWindEnergy = sum(AvgWindEnergy))
  
#Join to HUC_8 features
huc8_sf_join = merge(x = huc8_sf,
                     y = wind_raw,
                     by.x = 'HUC_8',
                     by.y = 'HUC_8')

#View the outputs
ggplot(data=huc8_sf_join) +
  geom_sf(aes(fill=SumWindEnergy)) +
  labs(
    title='Wind energy by HUC 8',
    fill ='MW/year'
  )
```

```{r Read in HUC12 watershed data}
#Read in uploaded data from Spatial Data folder
HUC12 <- st_read('./Data/Spatial/HUC12_Watershed_Boundaries/NHD_-_Watershed_Boundaries_HUC_12.shp')

#View using ggplot
ggplot(data=HUC12) + geom_sf()

#View using mapview
mapview(HUC12)

```


```{r Read in US Coal Field Data}
#Read in uploaded data from Spatial data folder
USCoalFields <- st_read('./Data/Spatial/US_Coal_Fields.shp')

#View using ggplot
ggplot(data=USCoalFields) + geom_sf()

#View using mapview
mapview(USCoalFields)

```


### 2.2.4 Reading in GeoJSON data

More and more spatial data are appearing on-line, in a format that allows us to connect directly to the data vs downloading local copies to our workspace. An example is the Homeland Infrstructure Foundation-Level Data (HIFLD) Their open data site (<https://hifld-geoplatform.opendata.arcgis.com/>) has links to many datasets.

When the data is served in GeoJSON format, we can ingest it directly in to R. Follow these steps: - Navigate to <https://hifld-geoplatform.opendata.arcgis.com/> - Scroll down to the Explore Categories area. Select Energy (for example) - Search for power plants. - Select the first return and open its [link](https://bit.ly/3aAVhz3) - Locate the APIs dropdown and copy the link to the GeoJSON option.

If you have difficulty, the link you want is: - <https://opendata.arcgis.com/datasets/ee0263bd105d41599be22d46107341c3_0.geojson>

This a link to a spatial dataset in GeoJSON format. Now let's read this dataset in and explore it:

```{r load.geojson.data}
powerplants_sf <- st_read('https://opendata.arcgis.com/datasets/ee0263bd105d41599be22d46107341c3_0.geojson')

#Reveal the field names
colnames(powerplants_sf)

#How many records
nrow(powerplants_sf)

#View on a map - specify the geometry column to draw faster
plot(powerplants_sf$geometry)

#Filter for just powerplants found in NC
nc_powerplants_sf <- powerplants_sf %>% 
  filter(STATE == "NC") 

#Have a look the variety of types (and number of each)
nc_powerplants_sf %>% 
  st_drop_geometry() %>% 
  count(TYPE, sort=TRUE)

#Most are solar farms; let's remove those
nc_powerplants_sf <- nc_powerplants_sf %>% 
  filter(TYPE != "SOLAR PHOTOVOLTAIC")

# Examine counts by fuel type
ggplot(nc_powerplants_sf) + 
  geom_sf(aes(color = PRIM_FUEL)) 

```
