---
output: 
  html_document: 
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Exploring Social Vulnerability, Power Plant Siting, and Wind Energy Potential in Wyoming"
subtitle: "https://github.com/jkrooney/Campbell_Rooney_ENV872_EDASp23_FinalProject"
author: "Sam Campbell & John Rooney"
fontsize: 12pt
mainfont: Times New Roman
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}

# Load your packages
library(tidyverse)
library(lubridate)
library(sf)
library(leaflet)
library(mapview); mapviewOptions(fgb = FALSE)
library(here)

#Disable on-the-fly projections
sf::sf_use_s2(FALSE)

# Set your working directory
setwd(here())

# Set your ggplot theme
my.theme <- theme_light(base_size = 14)+
  theme(axis.text = element_text(color = "grey19"), 
        legend.position = "top",
        legend.justification = "left")
theme_set(my.theme)

# Load your datasets
#read in counties data set
wy_counties <- st_read("./Data/Spatial/cb_2021_us_county_20m/cb_2021_us_county_20m.shp") %>% 
  filter(STATEFP == 56) #filter for WY

#read in svi data for WY
svi2020_wy_raw <- read.csv("./Data/Raw/Wyoming_COUNTY.csv", 
  colClasses = c('FIPS' = 'factor'))

#read in powerplant data for WY
powerplants_sf <- st_read('https://opendata.arcgis.com/datasets/ee0263bd105d41599be22d46107341c3_0.geojson')

#read in HUC12 dataset
HUC12 <- st_read("./Data/Spatial/HUC12_Watershed_Boundaries/NHD_-_Watershed_Boundaries_HUC_12.shp")

#Read in HUC12 wind energy data
wind_raw <- read.csv('https://raw.githubusercontent.com/ENV859/EnviroAtlasData/main/Wind_Energy.csv',
  colClasses = c('HUC_12' = 'factor'))
```

- Knitting commands in code chunks:
  -  `include = FALSE` - code is run, but neither code nor results appear in knitted file
  -  `echo = FALSE` - code not included in knitted file, but results are
  -  `eval = FALSE` - code is not run in the knitted file
  -  `message = FALSE` - messages do not appear in knitted file
  -  `warning = FALSE` - warnings do not appear...
  -  `fig.cap = "..."` - adds a caption to graphical results
  
  
# Rationale and Research Questions
The Inflation Reduction Act (IRA) was passed by the US Congress in late summer of 2022. The IRA is a generational piece of legislation providing funding and incentives for domestic energy production in the United States, offering the strongest government support yet for transitioning to renewable energy and confronting the challenges posed by climate change.

Among the many programs included in the IRA is one offering funding to help traditional fossil fuel communities and low-income communities transition to renewable energy production. To help us answer these questions, we decided to look at the state of Wyoming as a case study. Wyoming was selected for two main reasons. First, it is a state heavily reliant on traditional fossil fuel extraction and production. Second, we knew that Wyoming has relatively abundant wind energy potential, and that data existed that would help us understand and quantify wind energy potential throughout the state. 

This led us to two research questions:
  1. What relationship is there between socially vulnerable communities and the presence of power plants?   
  2. What potential is there for renewable energy production in communities considered socially vulnerable? 


\newpage

# Dataset Information
We worked with the following 5 datasets:

  1. County Boundary Data: In order to first map county lines onto our state of interest, we downloaded cartographic boundary shapefiles from the US Census Bureau website. The most recent year for these files was 2022. We used the 20m resolution file. Files were downloaded from the following URL: <https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html>.

  2. HUC12 Boundary Data: Because our wind energy data was at the HUC12 geographic level, we needed to download a cartographic boundary shapefile on HUC12s in Wyoming to then be able to assign HUC12 geography and wind energy figures to Wyoming counties. This would enable us to then do our comparative analysis with SVI and power plants at the county level. We downloaded this shapefile from the following URL: <https://water.geospatialhub.org/datasets/687dadd8b93f4ad092b64a84caf5b72a_0/explore?location=42.808013%2C-107.387472%2C6.66>.

  3. The Social Vulnerability Index (SVI) data was downloaded from the US Center for Disease Control Agency for Toxic Substances and Disease Registry website. The most recent year for which data is available is 2020. The geography selected was Wyoming, and the geography type was counties. We downloaded data as both shapefiles and csv file types, ultimately using the csv file. Data was pulled from the following URL: <https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html>. Additional documentation for the data was found here: <https://www.atsdr.cdc.gov/placeandhealth/svi/documentation/SVI_documentation_2020.html>.  

  4. Homeland Infrastructure Foundation-Level Data (HIFLD) for Power Plant types and locations in Wyoming was read directly into R using their open data site as GeoJSON data. The link used for this project was <https://opendata.arcgis.com/datasets/ee0263bd105d41599be22d46107341c3_0.geojson>. 

  5. Wind energy data estimated at the HUC12 level was read directly into R from the Environmental Protection Agency's EnviroAtlas data available online using the following URL: <https://raw.githubusercontent.com/ENV859/EnviroAtlasData/main/Wind_Energy.csv>. 

Our data wrangling included the following steps:

  With our SVI spatial data, we began by selecting for the columns of interest (County, FIPS, location, total population, number of people living below the poverty line, and number of minorities). We then used the ‘left_join()’ function to combine the SVI data with the County Boundary data. Next, we generated a new column in this combined data frame using the 'mutate' function to calculate the percentage of people living in poverty in each county, by dividing the number of people in poverty in each county by the total population of that county. We then changed the county name column from a character vector to a factor to enable us to generate a bar plot of poverty percentage by county. However, our code to generate a bar plot continued to generate errors, so we chose to instead use a scatterplot to visualize poverty percentage by county.  

  With our power plants spatial data, we began by filtering the data for power plants found in Wyoming. We then mapped this data with different colors to identify the different fuel types for each power plant. We overlayed these points on top of our SVI percentage data for the Wyoming counties, showing both a color gradient for percentage poverty and different colors for different power plant fuel types. The power plants data contained 13 different categories of fuel types, so we attempted to simplify our visualization by aggregating the power plants across major energy types (Fossil Fuel, Renewable, and Unknown). We attempted to use the ‘ifelse()’ function for this aggregation, but continually ran into errors; we think this may be because ‘ifelse()’ is only intended to create two groups, not three as we needed to.  

  With our HUC12 spatial data, we used the ‘st_centroid()’ function on the geometry column of the data set to generate centroids for each HUC12 in Wyoming. Our County Boundary data was using a different CRS than our HUC12 data (NAD83 for counties, and WGS 84 for HUC12s), so we used the ‘st_transform()’ function to change the CRS of our HUC12 data to match the CRS of the County Boundary data, using the EPSG code for NAD83 (4269). We then used the ‘st_join()’ function to combine our Counties data and HUC12 Centroids data, and to tag each HUC12 centroid with the county in which it falls.  

  With our HUC12 centroids generated and tagged with their corresponding counties, we were then able to use our HUC12-level wind data for analysis at the county level. We used the ‘left_join()’ function to combine our raw HUC12-level wind data with our wrangled data assigning HUC12s to counties, allowing us to see wind energy by county. We then tidied this combined data frame to select what columns we needed (county name, HUC12, average wind energy, and geometry). We used this data to generate a boxplot, showing the range of wind energy available in each county (the range includes the different average wind energy numbers for each HUC12 within a given county).  

  We then took an alternative approach to viewing wind energy availability, not as a range within a given county, but as an average for the whole county. We generated a new data frame in which we calculated the average wind energy for each county, and visualized this as a scatterplot to pair with our scatterplot of county-level poverty percentages. Similar to our poverty percentage data, we think a bar plot could have been a more effective visualization than a scatterplot, but ran into multiple errors when trying to generate a bar plot. 
  
The following table summarizes our data structure:

  

```{r svi data wrangling, include=FALSE}

#select needed columns
svi2020_wy_processed <- svi2020_wy_raw %>%
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV150, E_MINRTY)

#save processed file
write.csv(svi2020_wy_processed, row.names = F, file = "./Data/Processed/svi2020_wy_processed.csv")

```


```{r join svi to county data, include=FALSE}

#join svi to county data
counties_sf_join <- wy_counties %>% 
  left_join(svi2020_wy_processed, by = c("GEOID" = "FIPS") )

#Calculate percentage poverty by county with a new column
counties_sf_join <- 
  counties_sf_join %>% 
  mutate(PercentPov = ((E_POV150 / E_TOTPOP)*100))


#Change county name from a character vector to a factor
counties_sf_join$NAME <- factor(counties_sf_join$NAME)

```

```{r powerplant wrangling, include=FALSE}

#Filter for just powerplants found in WY
wy_powerplants_sf <- powerplants_sf %>% 
  filter(STATE == "WY") 

#Have a look the variety of types (and number of each)
wy_powerplants_sf %>% 
  st_drop_geometry() %>% 
  count(TYPE, sort=TRUE)

```

```{r compute centroids of HUC12s, include=FALSE}
#apply st_centroid function to geometry column of HUC12 dataset
HUC12Centroids <- HUC12 %>% 
  mutate(geometry = st_centroid(geometry))

#check that the HUC12 centroids show up as expected when mapped
mapview(HUC12Centroids)

#check the crs of each dataframe
st_crs(wy_counties)
#^uses NAD83
st_crs(HUC12Centroids)
#^uses WGS 84

#transform the crs for HUC12Centroids dataframe to match wy_counties dataframe
#EPSG code for NAD83 is 4269
HUC12CentroidsNAD83 <- 
  st_transform(HUC12Centroids,4269)

#check crs for transformed HUC12CentroidsNAD83 dataframe
st_crs(HUC12CentroidsNAD83)
#now uses NAD83, same as wy_counties dataframe

#Tag each centroid with the county in which it falls using st_join function
HUC12Counties <- 
  st_join(wy_counties, HUC12CentroidsNAD83) 

```

```{r wind data wrangling, include=FALSE}
#Join wind_raw dataframe with HUC12Counties dataframe to filter the wind data for Wyoming counties
HUC12CountiesWind <- HUC12Counties %>% 
  left_join(wind_raw, by = c("HUC12" = "HUC_12"))

#Create tidied dataframe of wind energy by county and HUC12
TidyHUC12CountiesWind <- HUC12CountiesWind %>% 
  select(NAME, NAMELSAD, STATE_NAME, HUC12, AvgWindEnergy, kWhkm2day, geometry)
```

```{r wrangling average wind by county, include=FALSE}
#Creating new data frame, calculating average wind energy by county
AvgWindByCounty <- 
  TidyHUC12CountiesWind %>% 
  drop_na(AvgWindEnergy) %>% 
  group_by(NAME) %>% 
  summarize(AvgWind = mean(AvgWindEnergy))
#^currently generates warning message "although coordinates are longitude/latitude, st_union assumes that they are planar"


```
\newpage

# Exploratory Analysis 
We used the ‘mapview()’ function to visualize the following:
  1. Wyoming counties and the number of people in poverty in each county

```{r spatial visualization of counties by poverty rate, echo=FALSE, fig.cap="Figure 1: Wyoming Counties"}
#View with mapview
mapview(counties_sf_join, 
        zcol = 'E_POV150')
```
 
  2. Wyoming HUC12s
  
```{r spatial visualization of HUC 12 data, echo=FALSE, fig.cap="Figure 2: HUC 12 Watersheds of Wyoming"}
#view HUC12 data with mapview
mapview(HUC12)
```

  3. The locations and fuel types of power plants in Wyoming

```{r powerplant visualizations, echo=FALSE, fig.cap="Figure 3: Location and Type of Power Plants in Wyoming"}
 
mapview(wy_powerplants_sf,
        zcol = 'TYPE', 
        position='bottomleft')
```

We additionally decided to explore the structure and contents of data frames as they were created, including our raw and processed SVI data.These can be seen in Figure 4. 
```{r structure SVI, echo=FALSE, fig.cap="Figure 4: Structure of Raw and Processed SVI Data Frames"}
#check raw structure
str(svi2020_wy_raw)

#Check processed structure
str(svi2020_wy_processed)
```

We additionally investigated the powerplants data before wrangling as seen in Figure 5. 
```{r powerplants explore, echo=FALSE, fig.cap="Figure 5: Exploration of Power Plants Data Frame"}
#Reveal the field names
colnames(powerplants_sf)

#How many records
nrow(powerplants_sf)

#structure of df
str(powerplants_sf)
```

\newpage

# Analysis
Our analysis built on our initial research questions and initial hypothesis, which were as follows: 
  Initial research questions:
    1. What correlation was there between socially vulnerable communities and the presence of power plants?
    2. What potential was there for renewable energy production in communities considered socially vulnerable? 

  Initial hypothesis: 
    Current areas of social vulnerability as judged by the CDC see higher rates of siting fossil fuel plants for electricity production. There is good potential to build on these overlapping conditions to make use of available wind energy resources for the transition to renewable energy as supported under the IRA.
    
We first generated a new column in our SVI dataframe to calculate the percentage of people in poverty in each Wyoming county, and visualized this using the 'mapview()' function to better see which counties were more impoverished relative to other counties (rather than viewing the number of people in poverty in each county as performed earlier, a visualization that was skewed by the differing populations of each county).

```{r percent poverty by county, echo=FALSE, fig.cap="Figure 6: Percentage of Population Living in Poverty"}
#View percentage poverty by county with mapview
mapview(counties_sf_join, 
        zcol = 'PercentPov')
```

We then used the 'mapview()' function to display a color gradient of poverty percentages within Wyoming counties on the same map as Wyoming power plants and their fuel types. The legend for power plant fuel types contains 13 types and so is quite large; we tried to simplify this into 3 major fuel types (Fossil Fuels, Renewables, and Unknown) using the 'ifelse()' function, but were unable to. By zooming out, we are able to view the whole map of Wyoming without being blocked by the legend of power plant fuel types.

```{r visualization of svi with powerplants, echo=FALSE, fig.cap="Figure 7: Visualization of Power Plant Locations and Population in Poverty By County"}
#map counties, svi, powerplants
mapview(counties_sf_join, 
        zcol = 'PercentPov') +
  mapview(wy_powerplants_sf,
          zcol = 'TYPE', 
          position = "bottomleft")
```

We calculated the centroids of our HUC12s to assign them to Wyoming counties, and then used the 'mapview()' function to visualize these HUC12 centroids across the state. 

```{r map of HUC12 by county, echo=FALSE, fig.cap="Figure 8: HUC 12 Watersheds Matched to Counties"}
#map newly created dataframe of HUC12Counties
mapview(HUC12Counties)
```

Once we had assigned the HUC12s to Wyoming counties, we were able to then match up our HUC12-level wind energy data with different counties. 

We first visualized the range of wind energy within each county by looking at the wind energy for different HUC12s within each county as a box and whisker plot.

```{r visualize the range of HUC12 wind energy within each county using a box and whisker plot, echo=FALSE, fig.cap="Figure 9: Wind Energy Potential by County"}
#view new tidied data as a boxplot
PlotWindByCounty <- 
  ggplot(TidyHUC12CountiesWind,
       aes(x = NAME,
           y = AvgWindEnergy)) +
  geom_boxplot() +
  labs(x = "County Name", y = "Avg Wind Speed in GWh / km^2 / day",
       title = "Wind Speeds by County in Wyoming") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(PlotWindByCounty)
```

We then developed a scatter plot to more succinctly visualize the average wind energy for each county. We think a bar plot might have enabled a clearer visualization of these average wind energy calculations, but this continually generated errors that we were unable to troubleshoot, so we chose to instead use a scatter plot.

```{r visualize average wind by county as scatter plot, echo=FALSE, fig.cap="Figure 10: Average Wind Energy Potential by County"}
#View new AvgWindByCounty data as a scatterplot
PlotAvgWindByCounty <- 
  ggplot(AvgWindByCounty,
       aes(x = NAME,
           y = AvgWind)) +
  geom_point() +
  labs(x = "County Name", y = "Avg Wind Speed in GWh / km^2 / day",
       title = "Average Wind Speed by County in Wyoming") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(PlotAvgWindByCounty)
```

To visualize county poverty levels in a similar, translatable format with county average wind energy levels, we then developed a scatter plot of poverty percentages at the county level. As with wind energy, we thought a bar plot might be a more ideal visualization, but ran into errors and insurmountable challenges with troubleshooting, so reverted to a scatter plot as our most practical option. 

We initially envisioned combining wind energy and poverty levels on a single map using the 'mapview()' function, but later determined that such a multi-level map would be more confusing than helpful for visualizing patterns in our data. 

We additionally envisioned using the 'facet_grid' function to display both scatterplots together, but when using the 'left_join' function to add average wind energy to our data frame including poverty information found that the new data frame inexplicably generated multiple average wind speeds for each county. These were listed under the correct wind speed as (for example) rows 1.1, 1.2, etc. We attempted to wrangle the new data frame to only include the correct wind speeds in rows 1, 2, 3, etc but could not separate them out. We decided that, for all practical purposes, the two scatterplots could still be shown one above the other even without the use of the 'facet_grid' function.

```{r scatterplot visualization poverty by county, echo=FALSE, fig.cap="Figure 11: Percentage of Population in Poverty by County"}
#View percentage poverty by county as a scatterplot
PlotSVICountyPercent <- 
  ggplot(counties_sf_join,
    aes(x = NAME,
        y = PercentPov)) +
  geom_point() +
  labs(x = "County Name", y = "% of Residents in Poverty",
       title = "Wyoming Poverty Levels by County") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(PlotSVICountyPercent)
```

## Question 1:  What correlation is there between socially vulnerable communities and the presence of power plants?

## Question 2: What potential is there for renewable energy production in communities considered socially vulnerable? 
In our visualization of county-level wind energy potential and county-level poverty rates, two counties that stood out as having high levels of both were Albany County and Carbon County. We also found that Albany County and Carbon County currently have more wind energy generation than most other Wyoming counties.  

Based on the data we currently have, this could be interpreted to mean that there is strong potential for socially vulnerable communities to transition away from fossil fuel-based electricity generation towards renewable energy generation. The type of renewable energy the county is able to transition to may then further depend on what renewable energy resources are most abundant locally; Albany County and Carbon County both have more wind energy than most other Wyoming counties.  

If communities are socially vulnerable and do not have good renewable energy potential (wind, solar, hydro, geothermal, etc.) locally available, they may encounter greater difficulty in transitioning away from fossil fuel-based electricity generation. As an alternative, they may be able to draw on transmission of renewable energy generation from other locations. The IRA’s support of renewable energy generation through tax credits will likely lower the costs of renewable energy generation in the coming years, and may ultimately make transmission of renewable energy from other locations a cheaper option than burning fossil fuels for local energy generation.  



\newpage

# Summary and Conclusions
Looking at the scatter plots, we find several counties that have relatively higher rates of residents living in poverty along with relatively high wind energy potential. They include Albany County, Carbon County, and Niobrara County. Albany County stands out as the county with both higher rates of poverty and higher wind energy potential. 

While Albany County already has some wind energy generation, it could be primed to expand. Albany County is home to the sixth largest population by county in the state, and the fourth largest city in Laramie, which is also home to the University of Wyoming. Casper, the second largest city in the state, is located catty-corner to Albany County to its northwest. Cheyenne, the most populous city in the state and capitol city, is located in Laramie County and neighbors Albany County directly to the east. Wind energy generated in Albany County could easily provide power to these population centers.  

\newpage

# References
The Inflation Reduction Act (2023). United States Environmental Protection Agency. https://www.epa.gov/green-power-markets/inflation-reduction-act 


